Taking Dyads Seriously

### 2017-10-29 paper/1_intro.tex
lil bits

We begin with a brief review of these dependencies and the AME model. Next, we conduct a simulation study to show how the AME approach can recover unbiased and well-calibrated regression coefficients in the presence of the dependencies that arise in dyadic data. Then, we apply this approach to five prominent studies in the international relations (IR) literature and compare results from the current state-of-the-art approach (a GLM with robust standard errors) to those obtained using the AME framework. The comparison reveals that in ignoring observational dependence, the standard approach overestimates the effect of key variables in the literature and underestimates their uncertainty. Consequently, the latent factor approach (AME) produces results that, at times, contradict those found in these studies in particular, and the broader literatures from which they are drawn. Moreover, we demonstrate the latent factor approach offers substantive insights that are often occluded by ignoring the interdependencies in most relational data that the field of IR is concerned with. Finally, we show that for each replication our network-based approach provides substantively more accurate out-of-sample predictions than the models used in the original studies. Thus, the AME approach can be used by scholars in the field to not only generate substantive insights, but it also enables us to better model the data generating process behind events of interest.  Most importantly, it facilitates concentration on the relations aspect of the field of international relations.

>>>>

We begin with a brief review of these dependencies and the AME model. Next, we conduct a simulation study to show how the AME approach can recover unbiased and well-calibrated regression coefficients in the presence of the dependencies that arise in dyadic data. Then, we apply this approach to five prominent studies in the international relations (IR) literature and compare results from the current state-of-the-art approach (a GLM with robust standard errors) to those obtained using the AME framework. The comparison reveals that in ignoring observational dependence, the standard approach overestimates the effect of key variables in the literature and underestimates their uncertainty. Consequently, the latent factor approach (AME) produces results that, at times, contradict those found in these studies in particular as well as the broader literatures from which they are drawn. Moreover, we demonstrate the latent factor approach offers substantive insights that are often occluded by ignoring the interdependencies in most relational data that the field of IR is concerned with. Finally, we show that for each replication our network-based approach provides substantively more accurate out-of-sample predictions than the models used in the original studies. Thus, the AME approach can be used by scholars in the field to not only generate substantive insights, but it also enables us to better model the data generating process behind events of interest.  Most importantly, it facilitates concentration on the relations aspect of the field of international relations.

### 2017-10-29 paper/3_sim.tex
lil bits

We utilize a simulation study to highlight the utility of AME as an inferential tool for dyadic analysis. The goal of the simulation is to assess how well AME can provide unbiased and well-calibrated estimates of coefficient parameters in the presence of unobserved dependencies. Specifically, we are concerned with conducting inference on regression parameters of a linear model for a network in the case where there is an omitted variable. Say that the true data-generating process for a particular $Y$ is given by:

@@

Last, beyond obtaining less biased and better-calibrated parameter estimates, a key benefit of the AME framework is that we can also estimate unobserved dependencies through the random effects structure of the model. In the case of the data generating process for $Y$, $W$ is set as an unobserved dyadic covariate that had a homophilous effect on $Y$. Homophilous because $W$ within this framework is simply an example of a dyadic attribute involving $i$ and $j$ that positively affects the degree to which they will interact with one another, i.e., $y_{ij}$. This type of unobserved dependency will be captured through the multiplicative effects portion of the model, $\mathbf{U}^{\top} \mathbf{D} \mathbf{V}$. To estimate how well the model performs in doing this we recover the multiplicative effects term for each simulation and calculate the correlation between it and the unobserved dependency, $W$.\footnote{Specifically, since both the multiplicative effects term and $W$ are continuous dyadic variables, we just calculate the Pearson correlation coefficient.} We visualize the distribution of the correlations from each of the 1,000 simulations in Figure~\ref{fig:ameCorr} for the case where the number of actors was set to 100 (top pair of panels) and 50. Additionally, we calculate the median across the correlations and display that result using a vertical line. For both $n=50$ and $n=100$, we find that the multiplicative effects perform very well in capturing the unobserved dependency, which indicates that the structure provided by this framework is not simply capturing noise but can be used as a tool to estimate unobserved structure.

>>>>

We utilize a simulation study to highlight the utility of AME as an inferential tool for dyadic analysis. The goal of the simulation is to assess how well AME can provide unbiased and well-calibrated estimates of coefficient parameters in the presence of unobserved dependencies. Specifically, we are concerned with conducting inference on regression parameters of a linear model for a network in the case where there is an omitted variable. For instance, assume that the true data-generating process for a particular $Y$ is given by:

@@

Last, beyond obtaining less biased and better-calibrated parameter estimates, a key benefit of the AME framework is that we can also estimate unobserved dependencies through the random effects structure of the model. In the case of the data generating process for $Y$, $W$ is set as an unobserved dyadic covariate that had a homophilous effect on $Y$. Homophilous because $W$ within this framework is simply an example of a dyadic attribute involving $i$ and $j$ that positively affects the degree to which they will interact with one another, i.e., $y_{ij}$. This type of unobserved dependency will be captured through the multiplicative effects portion of the model, $\mathbf{U}^{\top} \mathbf{D} \mathbf{V}$. To estimate how well the model performs in doing this we recover the multiplicative effects term for each simulation and calculate the correlation between it and the unobserved dependency, $W$.\footnote{Specifically, since both the multiplicative effects term and $W$ are continuous dyadic variables, we calculate the Pearson correlation coefficient.} We visualize the distribution of the correlations from each of the 1,000 simulations in Figure~\ref{fig:ameCorr} for the case where the number of actors is set to 100 (top pair of panels) and 50. Additionally, we calculate the median across the correlations and display the result using a vertical line. For both $n=50$ and $n=100$, we find that the multiplicative effects perform very well in capturing the unobserved dependency, which indicates that the structure provided by this framework is not simply capturing noise but can be used as a tool to estimate unobserved structure.

### 2017-10-29 paper/4_app.tex
lil bits

We obtained the data for each of these studies from their replication archives and replicated the main results of each article.\footnote{Without exception this was straightforward to accomplish, thanks to an increasing norm in the social sciences of open data sharing.} The broader goal in doing this, beyond introducing the use of the AME framework in an applied setting, is to examine the extent to which interdependencies within typical dyadic data make much difference in what we have learned about international relations from recent empirical studies using a framework that a priori assumes conditional independence.

% SM: maybe too on the nose (Whether putative control variables, for example, are precisely estimated is less important.) 
Typically, scholars have a small set of independent variable in a complicated empirical model that they look at for evaluating the validity of their empirically estimated models. For example, in the study by Reiter and Stam (2003) the goal is to determine whether in mixed dyads---consisting of a democratic country and country ruled by a personalist dictatorship---it is the democratic country or the personalist country that is associated with a greater likelihood of the dyad being in a dispute.  In a regression including fourteen covariates, they focus on whether the democratic nature of the initiating country is a statistically significant covariate.\footnote{In particular, the variables \texttt{Pers/Democ} and \texttt{Democ/Pers} in Model 3, in Table 1, page 335.} In our re-examination of Reiter and Stam, we focus on whether the model replicates when re-estimated within the AME framework. Each of the other four studies also has a crucial finding that we hone in on to further draw into focus the potential differences brought about by utilizing an AME estimation procedure.  In Table~\ref{tab:modelFindingSumm}, we present the overall results; the term \textit{does not replicate} indicates only that the sign and/or significance of the putatively crucial finding in the original study is not found to hold in the AME estimation.\footnote{Full tabular results for each of the original and reestimated models are presented in the Appendix.}

@@

\citet{reiter:stam:2003} examine the relationship between democracy, dictatorship and the initiation of militarized disputes.  They use directed dyads and find that dyads involving a democratic leader on the one hand and a personalist dictator on the other tend to be violent. They also discover that dictators are likely to challenge democracies, but that the converse is not true.  In addition, military regimes and single-party regimes are more prone to initiate disputes with democracies, than the other way around. Independent variables are largely taken from an earlier study and focus on various encodings of regime types, contiguity, alliance, and capability measures. As is prevalent in these kinds of studies, Reiter \& Stam employ a logistic regression that includes an indicator of the time since the last dispute as well as three cubic splines. The database for this study is constructed using \texttt{EUGene} \citep{bennett:stam:2000} and comprises approximately three-quarters of a million stacked dyads. Based on their statistical analysis, they conclude that institutional constraints affect the propensity of democratic and non-democratic leaders to engage in military conflict.

>>>>

We obtained the data for each of these studies from their replication archives and replicated the main results of each article.\footnote{Without exception this was straightforward to accomplish, thanks to an increasing norm in the social sciences of open data sharing.} The broader goal in doing this, beyond introducing the use of the AME framework in an applied setting, is to examine the extent to which interdependencies within typical dyadic data alter previous conclusions about international relations from recent empirical studies using a framework that a priori assumes conditional independence.

% SM: maybe too on the nose (Whether putative control variables, for example, are precisely estimated is less important.) 
Typically, scholars are concerned with conducting hypothesis testing on a small set of independent variables that they have assembled in a complicated empirical model. For example, in the study by Reiter and Stam (2003) the goal is to determine whether in mixed dyads---consisting of a democratic country and country ruled by a personalist dictatorship---it is the democratic country or the personalist country that is associated with a greater likelihood of a dispute dyad being in a dispute.  In a regression including fourteen covariates, they focus on whether the democratic nature of the initiating country is a statistically significant covariate.\footnote{In particular, the variables \texttt{Pers/Democ} and \texttt{Democ/Pers} in Model 3, in Table 1, page 335.} In our re-examination of Reiter and Stam, we focus on whether the model replicates when re-estimated within the AME framework. Each of the other four studies also has a crucial finding that we hone in on to further draw into focus the potential differences brought about by utilizing an AME estimation procedure.  In Table~\ref{tab:modelFindingSumm}, we present the overall results; the term \textit{does not replicate} indicates only that the sign and/or significance of the putatively crucial finding in the original study is not found to hold in the AME estimation.\footnote{Full tabular results for each of the original and reestimated models are presented in the Appendix.}

@@

\citet{reiter:stam:2003} examine the relationship between democracy, dictatorship and the initiation of militarized disputes.  They use directed dyads and find that dyads involving a democratic leader on the one hand and a personalist dictator on the other tend to be violent. They also discover that dictators are likely to challenge democracies, but that the converse is not true.  In addition, military regimes and single-party regimes are more prone to initiate disputes with democracies, than the other way around. Independent variables are largely taken from an earlier study and focus on various encodings of regime types, contiguity, alliance, and capability measures. As is prevalent in this literature, Reiter \& Stam employ a logistic regression that includes an indicator of the time since the last dispute as well as three cubic splines. The database for this study is constructed using \texttt{EUGene} \citep{bennett:stam:2000} and comprises approximately three-quarters of a million stacked dyads. Based on their statistical analysis, they conclude that institutional constraints affect the propensity of democratic and non-democratic leaders to engage in military conflict.

### 2017-10-29 paper/5_concl.tex
lil bits

To explore this approach in the context of international relations we have presented two broad analyses. The first is a simulation where the characteristics of the network are known. This shows that, when there are unobserved dependencies, the AME approach is less biased in terms of parameter estimation compared with standard approach employed in international relations to study dyadic data (i.e., GLM models). The second is a replication of five prominent studies that have been published recently using a broad range of dyadic data to draw inferences about international relations.  These five studies have been replicated with the original research designs, each of which used a statistical method that assumes the dyadic data are all independent from one another.  We then re-analyzed each study using the AME model.  In every case, we found that the AME approach provided a) increased precision of estimation, b) better out-of-sample fits, c) evidence of 1st-, 2nd-, and 3rd-order dependencies that were overlooked in the original studies.\footnote{The Appendix contains performance data on all of these replications, as well as sample code illustrating how to undertake AME analysis using \texttt{amen}.} In several cases, the new approach overturns the basic findings of the original research.  This leads us to speculate that many of the findings in the international relations literature may be fragile in the sense that they can only be obtained under stringent assumptions that cannot possibly be valid.  In turn, this leads to a certain arbitrariness in some research findings, which might lead to puzzles that are more apparent than real \citep{zinnes:1980}.

>>>>

To explore this approach in the context of international relations we have presented two broad analyses. The first is a simulation where the characteristics of the network are known. This shows that, when there are unobserved dependencies, the AME approach is less biased in terms of parameter estimation compared with standard approach employed in international relations to study dyadic data (i.e., GLM models). The second is a replication of five prominent studies that have been published recently using a broad range of dyadic data to draw inferences about international relations.  These five studies have been replicated with the original research designs, each of which used a statistical method that assumes the dyadic data are all independent from one another.  We then re-analyzed each study using the AME model.  In every case, we found that the AME approach provided a) increased precision of estimation, b) better out-of-sample fit, c) evidence of 1st-, 2nd-, and 3rd-order dependencies that were overlooked in the original studies.\footnote{The Appendix contains performance data on all of these replications, as well as sample code illustrating how to undertake AME analysis using \texttt{amen}.} In several cases, the new approach overturns the basic findings of the original research.  This leads us to speculate that many of the findings in the international relations literature may be fragile in the sense that they can only be obtained under stringent assumptions that cannot possibly be valid.  In turn, this creates a certain arbitrariness in some research findings, which might lead to puzzles that are more apparent than real \citep{zinnes:1980}.
